A semantics of core AUTOSAR

### 2016-09-02 paper/paper.tex
Minor fixups.

In common with most real-time process calculi, the concept of passing time in our calculus is kept separate from the computational work expressed by the say/hear transitions \cite{Baeten00processalgebra}. This has the advantage of freeing the semantics from possessing a particular computation speed, as any finite number of computation steps can be performed before time must advance. On the other hand, it also means that the  limitations of a particular platform cannot be directly studied unless the calculus is complemented with some form of constraint on the work-time relationship. Such an extension is beyond the scope of the current report, though.

@@

The specification restricts client-server connections such that a client may not be connected to multiple servers, for the purpose of avoiding an operation call to be handled by more than one server runnable \cite[ch.~4.2.3]{AR:RTE}. However, this restriction does not explicitly forbid multiple \emph{Operation\-InvokedEvent}s referring to the same client-server operation to trigger separate runnables. Our formulation of client-server communication is based on the assumption that this is just an oversight, as allowing multiple events on a server operation would have the same problem with conflicting response values as a would one-to-many client-server connection pattern. It is actually not easy to see what could constitute an alternative interpretation here, without introducing a need for some new programming mechanism for selecting among server results.

@@

A similar specification ambiguity concerns \emph{DataReceived\-Event}s in the presence of a full receiver buffer. We have chosen to block these events from activating connected runnables (axiom (30)), although the opposite choice could also be justified on grounds that activation of a receiver must be even more pressing when its buffer is overflowing. Expressing this formally would amount to simply deleting axiom (30) together with the last precondition of axiom (29).

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in our simplistic proposal. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative resolution of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its environment. We may for example wish to define that a special error code is returned if our runnable is running on behalf of a client, and that any exclusive areas held are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example involving an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by a formal semantics approach.

@@

All aspects of the AUTOSAR specification that concern representation and selection of data values are beyond the scope of the present study. This includes
\begin{itemize}
\item Data types on any level of abstraction, and their associated mappings \cite[ch.~5]{AR:SWC}.
\item Port interface mapping and data scaling \cite[ch.~4.3]{AR:SWC}.
\item Measurements and calibration parameters \cite[ch.~2.2]{AR:SWC}.
\item Variant handling \cite[ch.~2.4]{AR:SWC}.
\end{itemize}
These modeling elements only impose restrictions and demands on the values $v$ and continuations $k$ we reference in our semantics, and which we in turn do not constrain in any way.

@@

The only atomic software component type we include on our formalization is the \emph{Application} software component type. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuator}, \emph{ComplexDeviceDriver}, \emph{Service}, \emph{ServiceProxy}), or trivial in comparison (\emph{NvBlock}, \emph{EcuAbstraction}, \emph{Parameter}) \cite[ch.~3.2.3]{AR:SWC}.

>>>>

In common with most real-time process calculi, the concept of passing time in our calculus is kept separate from the computational work expressed by the say/hear transitions \cite{Baeten00processalgebra}. This has the advantage of freeing the semantics from possessing a particular computation speed, as any finite number of computation steps can be performed before time must advance. On the other hand, it also means that the  limitations of a particular platform cannot be directly studied unless the calculus is complemented with some form of constraint on the work/time relationship. Such an extension is beyond the scope of the current report, though.

@@

The specification restricts client-server connections such that a client may not be connected to multiple servers, for the purpose of avoiding an operation call to be handled by more than one server runnable \cite[ch.~4.2.3]{AR:RTE}. However, this restriction does not explicitly forbid multiple \emph{Operation\-InvokedEvent}s referring to the same client-server operation to trigger separate runnables. Our formulation of client-server communication is based on the assumption that this is just an oversight, as allowing multiple events on a server operation would have the same problem with conflicting response values as would a one-to-many client-server connection pattern. It is actually not easy to see what could constitute an alternative interpretation here, without introducing a need for some new programming mechanism for selecting among server results.

@@

A similar specification ambiguity concerns \emph{DataReceived\-Event}s in the presence of a full receiver buffer. We have chosen to block these events from activating connected runnables (axiom (30)), although the opposite choice could also be justified on grounds that activation of a receiver might be even more pressing when its buffer is overflowing. Expressing this formally would amount to simply deleting axiom (30) together with the last precondition of axiom (29).

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in our simplistic proposal. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative resolution of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its environment. We may for example wish to define that a special error code is returned if our runnable is a server running on behalf of a client, and that any held exclusive areas are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example involving an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by a formal semantics approach.

@@

All aspects of the AUTOSAR specification that concern representation and selection of data values are beyond the scope of the present study. This includes
\begin{itemize}
\item Data types on any level of abstraction, and their associated mappings \cite[ch.~5]{AR:SWC}.
\item Port interface mapping and data scaling \cite[ch.~4.3]{AR:SWC}.
\item Measurements and calibration parameters \cite[ch.~2.2]{AR:SWC}.
\item Variant handling \cite[ch.~2.4]{AR:SWC}.
\end{itemize}
These modeling elements only impose restrictions and demands on the values $v$ and continuations $k$ we reference in our semantics, and which we, in our turn, do not constrain or interpret in any way.

@@

The only atomic software component type we include in our formalization is the \emph{Application} software component type. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuator}, \emph{ComplexDeviceDriver}, \emph{Service}, \emph{ServiceProxy}), or trivial in comparison (\emph{NvBlock}, \emph{EcuAbstraction}, \emph{Parameter}) \cite[ch.~3.2.3]{AR:SWC}.
